---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# when

<!-- badges: start -->
<!-- badges: end -->

The *When* dimension plays a fundamental role in *Multidimensional Systems*, it allows us to express **when** the analysed focus of attention have occurred.

The purpose of the `when` package is to assist in the implementation of the *When* dimension in multidimensional systems. In particular, it supports the generation of tables for the implementation of **date** and **time** in ROLAP (*Relational On-Line Analytical Processing*) systems.

## Installation

You can install the development version of `when` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("josesamos/when")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(when)
## basic example code
```

