---
title: "Generating the *When Dimension* based on date and time tables"
date: "2023-12-26"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{Generating the *When Dimension* based on date and time tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The *When Dimension* plays a fundamental role in *Multidimensional Systems*, it allows us to express **when** the analysed focus of attention have occurred.

This dimension corresponds to the generically called “Time Dimension”, as named in @kimball1996data and @adamson2010star. Later it has also been called "Date Dimension" or "Time Dimension" to express granularity [@kimball2013data]. We prefer to call it "When Dimension" because granularity is not involved in that term.

Although conceptually the date and time can be represented together in a single dimension, in ROLAP (*Relational On-Line Analytical Processing*) systems, it is common to implement these concepts separately, sometimes in separate tables "to avoid a row count explosion in the date dimension" [@kimball2013data] by adding the time for each day.

In @kimball2013data these dimensions are considered *static dimensions*: they are built at the beginning of the project for the period under consideration and are referenced when adding the instances to the fact table. As mentioned there: "Typically, these dimensions are built in an afternoon with a spreadsheet."

The purpose of the `when` package is to assist in the implementation of the ***When Dimension*** in Multidimensional Systems implemented on a ROLAP star database, regardless of its granularity, with the following features:

- It supports the generation of tables with the granularity needed (second, minute, hour, date, week, month or year). 

- We consider separate tables for **time** data and data obtained from **date** to avoid a row count explosion if a single table were used. 

- It can be used to generate tables **statically**, indicating a period, or **dynamically**, indicating data series. 

- In addition to attributes at the selected granularity, other attributes can be added at coarser granularity to facilitate querying.

The rest of this document is structured as follows: First, the general process of defining these dimensions is presented. Next, a section is dedicated to showing examples of dimension definitions, time-based and date-based. Finally, the document ends with conclusions and bibliography.

# Definition process

The definition process consists of the following steps:

1. Creating an object.
1. Configuration of general aspects.
1. Definition of instances.
1. Selection and definition of levels.
1. Generation of the result.
1. Table export.

Steps 2, 3 and 4 can be done in any order, they can even be included in step 1.

Below we discuss each of them.

## Creating an object

The definition process begins by creating an object of the `when` class.

```{r setup}
library(when)

w_date <- when()
```

By default, an object is created based on date, but we can also create it based on time by indicating it using the `type` parameter.

```{r}
w_time <- when(type = 'time')
```

Virtually all configuration options can be defined using the object creation function. They can also be defined later using specific functions, as shown in the following section.

## Configuration of general aspects

For example, using the following combination of functions we obtain the same result as defining the type at the time of object creation.

```{r}
w_time_2 <- when() |>
  configure_dimension(type = 'time')

identical(w_time, w_time_2)
```

By default it considers the system language. Since our system is Spanish, for the literals of day and month names to appear in English, we must configure it explicitly.

```{r}
w_date <- w_date |>
  configure_dimension(locale = Sys.setlocale("LC_TIME", "English"))
```

By default the dimension table will include a surrogate key. Through the `surrogate_key` parameter of this function we can indicate that it is not included.

Next we have to select and configure the levels and define the instances. These operations can be carried out in any order.

## Definition of instances

We can define the instances by giving a range of values or by indicating the specific instances. The values in both cases will depend on the type of dimension (date-based or time-based).

We can indicate these values at the time of creating the class object or through the `define_instances()` configuration function.

- In the case of date, we can indicate a year, year and month (with the format `yyyy-mm`), or year month and day (`yyyy-mm-dd`).

- In the case of time, we can indicate an hour, hour and minute (with the format `hh:mm`), or hour, minute and second (`hh:mm:ss`).

In both cases we can use the string format, the one provided by the [`lubridate`](https://CRAN.R-project.org/package=lubridate) package or, when it is a single number (year or hour) we can use an integer value. Then it will be transformed appropriately.

For example, below they are used with the format of `lubridate`.

```{r}
w_date <- w_date |>
  define_instances(start = lubridate::today(),
                   end = lubridate::today() + lubridate::years(5))
```

The following definition will be appropriate if we want to obtain the dimension at the year level, because we are indicating the specific values to consider (not a period).

```{r}
w_date_2 <- w_date |>
  define_instances(values = 2020:2030)
```

The following definition will be appropriate at any level of detail because we are giving the range of values to consider (dates between those two years).

```{r}
w_date_3 <- w_date |>
  define_instances(start = 2020, end = 2030)
```

When only the year is indicated, it is considered its first day. Therefore the previous definition is equivalent to the following one.

```{r}
w_date_4 <- w_date |>
  define_instances(start = "2020-01-01", end = "2030-01-01")

identical(w_date_3, w_date_4)
```

In the case of time, by default, it is considered at second level and all seconds of the day.

```{r}
w_time_3 <- w_time |>
  define_instances(start = "00:00:00", end = "23:59:59")

identical(w_time, w_time_3)
```


## Selection and definition of levels

In the definition, by default, the attributes that we have considered to be most frequently used have been included. We can consult them for each level using the `get_level_attribute_names()` function with the parameter `selected = TRUE`. If we define the parameter as `selected = FALSE` (default option), it will return all available attributes. We can indicate the name of the level or, if we do not indicate any, it will obtain the attributes of all the levels. We obtain the level names using the `get_level_names()` function. Below are examples of using these functions.

```{r}
w_date |>
  get_level_attribute_names(selected = TRUE)

w_date |>
  get_level_names()

w_date |>
  get_level_attribute_names(name = 'month', selected = TRUE)

w_date |>
  get_level_attribute_names(name = 'month')

w_time |>
  get_level_attribute_names()
```

There is a function to configure each level: the attributes defined by default are configured there. We can select or deselect each attribute individually using the attribute name as a boolean parameter of the function. For the levels associated with the date, we can also select all attributes defined at the level using the special `include_all` parameter.

For example, if we do not want to include the name of the month, we can configure it using the `include_month_level()` function or at the time of object creation with a parameter of the same name, as shown below.

```{r}
w_date_5 <- w_date |>
  include_month_level(month_name = FALSE)

w_date_6 <- when(
  start = lubridate::today(),
  end = lubridate::today() + lubridate::years(5),
  month_name = FALSE
)

identical(w_date_5, w_date_6)

w_date_5 |>
  get_level_attribute_names(name = 'month', selected = TRUE)
```

There is available a similar function for each level: `include_time_level()`, `include_day_level()`, `include_week_level()`, `include_month_level()` and `include_year_level()`.

For date-based levels, we can include or exclude them using the `level` parameter of each of those functions or using the `include_date_levels()` function. The advantage of this last function is that we can include or exclude several levels in a single step.

```{r}
w_date_7 <- w_date |>
  include_month_level(level = FALSE)

w_date_8 <- w_date |>
  include_date_levels(month_level = FALSE)

identical(w_date_7, w_date_8)

w_date_7 |>
  get_level_attribute_names(name = 'month', selected = TRUE)
```

If we want to include the level with the default attributes, just use the `include_month_level()` function again with the default values of the parameters.

```{r}
w_date_9 <- w_date_7 |>
  include_month_level()

w_date_9 |>
  get_level_attribute_names(name = 'month', selected = TRUE)
```


## Generation of the result

Once the characteristics of the dimension are defined, we can generate its table.

Although we have not yet generated the dimension table, the following function returns the names of the attributes it will contain. They can be obtained as a string or as a vector of strings.

```{r}
w_date |>
  get_table_attribute_names(as_string = FALSE)
```

If it is generated in the form of a string (`as_string = TRUE`, default option), the result can be used to easily change the name of the attributes, when the table is generated, using the `set_table_attribute_names()` function, if deemed necessary.

When we consider that the dimension is already properly defined, we can generate its corresponding table using the `generate_table()` function. We update the definition of the object. Any errors found will be notified to us.

```{r}
w_date <- w_date |>
  generate_table()

w_time <- w_time |>
  generate_table()
```

Once the table is generated, we can export it. If we want to access the table and it is not yet generated, it will send us an error message.

If we make changes to the configuration, it is our responsibility to generate the table again.

## Table export

We can access the dimension table to work with it in R (for example using the [`rolap`](https://CRAN.R-project.org/package=rolap) package), or export it in the format that is most convenient for us to work with other tools.

Next we access the generated tables and display them. First of all for the date.

```{r}
t_date <- w_date |>
  get_table()
```

The first and last rows of the obtained result are shown below.

```{r, results = "asis"}
pander::pandoc.table(rbind(head(t_date, 5), tail(t_date, 5)),
                     split.table = Inf)
```

Next for the time.

```{r}
t_time <- w_time |>
  get_table()
```

The start and end of the result table is shown below.

```{r, results = "asis"}
pander::pandoc.table(rbind(head(t_time, 5), tail(t_time, 5)),
                     split.table = Inf)
```

`day_part` is predefined in English. Literals or associated hours can be accessed and changed using the `get_day_part()` and `set_day_part()` functions.

We can export the generated tables to any Relational DBMS, as shown below, we just have to pass a connection to the database as a parameter..

```{r}
my_db <- DBI::dbConnect(RSQLite::SQLite())

w_date |>
  get_table_rdb(my_db)

w_time |>
  get_table_rdb(my_db)

DBI::dbListTables(my_db)

DBI::dbDisconnect(my_db)
```

In the previous example, since we have not explicitly defined any names for the tables in the configuration, they have been assigned as a name the element on which they are based (date and time).

Functions to store the tables in *xlsx* and *csv* format are also available: `get_table_xlsx()` and `get_table_csv()`.


# Definition examples

## Time based dimension

## Date based dimension

# Conclusions

`when` package offers a set of operations that allow us to transform tables into star databases.  From several star databases you can define a constellation with conformed dimensions.

Operations have been designed to be intuitive and easy to use. The result greatly facilitates the data transformation process for the exposed situation.

The implementation of the multidimensional database obtained can be exported to work with multidimensional analysis tools on spreadsheets or RDBMS.

# Bibliography

